Data Management \\ and LSST Special Programs

### 2018-06-18 DMTN-065.tex
Set repo location
major revisions; issued

\documentclass[DM,lsstdraft,toc]{lsstdoc}
\usepackage{graphicx}
\usepackage{url}
\usepackage{latexsym}
\usepackage{color}
% black, blue, brown, cyan, darkgray, gray, green, lightgray, lime, magenta, blue, orange, pink, purple, red, teal, violet, white, yellow.
\usepackage{enumitem}

@@

\setDocRef{DMTN-065}
\date{\today}
\setDocRevision{TBD}
\setDocStatus{draft}

\setDocAbstract{\textbf{This is a draft document and may yet change.} The purpose of this study is to ensure that the LSST Data Management's plans for processing data from Special Programs (Deep Drilling Fields and Mini-Surveys) will meet the needs of the scientific community, and that these plans have been appropriately sized for computational processing and are accurately represented in the work-hours budget. Unresolved issues spawn \textbf{Actions}, which are typically documentation change requests or proposed studies to assess a potential problem.}

\setDocChangeRecord{%
\addtohist{1}{2017-11-??}{Internal working document.}{Melissa Graham}
%\addtohist{2}{yyyy-mm-dd}{Future changes}{Future person}
}

@@

The intent of the LSST Data Management team for processing data from Special Programs can be summarized by the following statement, which reflects Section 6 of the Data Products Definitions Document (\DPDD, \citeds{LSE-163}; but see also the minor proposed change to this statement in \ref{Size-1}):

\begin{quote}
LSST will not write unique algorithms for processing Special Programs data or reprocessing Main Survey data, but LSST will reconfigure the pipelines and generate imaging and catalog products for Special Programs data whenever possible, and LSST will make its codes accessible to the science community and commit $\sim$10\% of its computing resources toward enabling Level 3 analysis and data product creation, including user-driven Special Programs processing.
\end{quote}

\noindent The purpose of this study is to ensure that the requirements and plans of DM are accurately expressed in all major DM documentation, and that the computational system and human work-hours needed to accomplish Special Programs processing have been accurately evaluated. This study spawns \textbf{Action} items, which are typically either documentation change requests or proposals for future studies needed to resolve an issue.

%In particular, this study is designed to review: (1) the processing and products that are required to enable science for Special Programs, (2) the current plans for DM processing and products with respect to Special Programs, (3) whether the DM plans meet the expected needs of Special Programs, and (4) any necessary changes to be written into the requirements, designs, and plans, in which case an \textbf{Action} is spawned.

Section \ref{sec:data} compiles an overview of the potential diversity in data from Special Programs, compared to the Wide-Fast-Deep (WFD) main survey data: standard visit images of $2\times15$ seconds for a wide area of the nighttime sky, obtained with a dither pattern using up to five filters. We assemble the possible limitations on Special Programs observations from the capabilities of the telescope and camera.

Section \ref{sec:dmplans} reviews the current DM plans for processing Special Programs data, including the capabilities for processing non-standard visits and whether DM might need to impose boundaries on the parameters of non-standard images that its pipelines will be able to processes. We also describe DM's intent for incorporating Special Programs data into the WFD main survey products, reconfiguring pipelines to create unique products, and providing support for end-user Level 3 processing.

Section \ref{sec:docrev} steps through each of the relevant LSST documents and highlights specific requirements or text in these documents that might need additions or clarifications to properly reflect DM's plans for processing Special Programs data.
%These \textbf{Concerns} are not prioritized by urgency or importance, but instead represent more of a complete set of possible issues. \textbf{The Concerns raised by this study fall into two main categories:} \\
%1. Whether DM will be able to process the full diversity of non-standard visit images, and if not, what the boundaries will be. \\
%2. Ensuring that DM's intent to process Special Programs data with reconfigured pipelines, and to provide separate data products that meet the science goals of each program, is accurately reflected in all documents, as it is in the DPDD, and is in the staffing budget. \\

Section \ref{sec:SP_SC} gathers the existing efforts from the Science Community regarding Special Programs, summarizes the scientific needs of as many currently proposed Special Programs as we can find, and outlines the future mechanisms for LSST to gather more community input -- including some recommendations for the future call for Special Programs white paper proposals.

Section \ref{sec:SPCS} contains a series of case studies for DM's processing of Special Programs data, which are based on several of the currently circulating ideas for deep drilling fields and mini-surveys. These processing case studies contribute to this work by offering an independent vantage point from which to identify a few additional issues regarding the processing of Special Programs data, and could also serve as examples for the upcoming call for community white papers, which will request that a similar section be included in all proposals.

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
\clearpage
\section{The Potential Diversity of Special Programs Data} \label{sec:data}

To better identify what might be missing in the DM plan to accommodate Special Programs, we need a comprehensive understanding of the potential diversity in future Special Programs data. The concern is that if Special Programs data is significantly different from the Wide-Fast-Deep survey of tiled, nighttime, 15--30 second exposures (i.e., standard visit images), then it might not be adequately handled by the DM system. To get an idea of the potential diversity of Special Programs data, we first attempt to understand the technical limitations that the facility and instrumentation places on the observing modes, e.g., exposure times, filter changes, in Section \ref{ssec:data_bounds}.

We can also get an idea of the expected diversity in exposure type and survey pattern from the current set of proposed Special Programs for LSST, which are reviewed in more detail in Section \ref{sec:SP_SC}. We find that most of the proposed Special Programs are likely to use standard visits, two $\sim15$ second exposures, and that the potential diversity of Special Programs data is mainly comprised of exposures that are significantly shorter than the WFD main survey standard visit exposures of $2\times15$ seconds, exposures that are obtained with a bright sky background during twilight, long series of exposures obtained of the same field (i.e., deep drilling), and images of crowded fields.

All aspects related to DM's \textit{processing} of Special Programs data are addressed in Section \ref{sec:dmplans}.

% % % % % % % % % % % % % % % % % %
\subsection{Technical Boundaries for Special Programs Data}\label{ssec:data_bounds}

We can constrain the potential diversity of Special Programs data if we know the observing boundaries that will be imposed on the accepted proposals. For the purposes of this document we consider the ``observing boundaries" as those imposed by the camera, telescope, or site, and independent of any DM processing capabilities. It is not the job of this document to define these observing boundaries, but we do need to know what they are in order to predict the diversity of incoming data that DM will be processing.

The current technical limitations on what kind of observations can be obtained with the LSST telescope and camera are described below in a rough draft, to be updated in the future. Additionally, in Section \ref{ssec:docrev_oss}, we review the Observatory System Specifications (OSS, \citeds{LSE-30}) document and highlight a couple more questions regarding potential boundaries on Special Programs observations.

Filter Changes: \\
$\bullet$ The maximum time for filter change is 120 seconds: 30 seconds for the telescope to reorient the camera to its nominal zero angle position on the rotator, and 90 seconds to the camera subsystem for executing the change (OSS-REQ-0293). \\
$\bullet$ Note that changes regarding the requirements on filters in support of "deep drilling fields" are underway here: \href{https://project.lsst.org/groups/ccb/node/938}{LCR-458}. \\
$\bullet$ \textit{(MLG - The next three points are from a document-in-progress by Ivezic and Ritz; update to formal citation when it's public-facing.)} \\
$\bullet$ The minimum time between filter changes has no restrictions from e.g., thermal tolerances. However, based on overheads and efficiency, it is recommended to keep the filter change rate lower than once every 10 minutes. \\
$\bullet$ The maximum total number of filter changes is 100,000 over 15 years, an average of 18 changes per night. \\
$\bullet$ The maximum number of filter swaps in/out of the carousel is 3000 in 15 years, or once every two nights. \\

Exposure Times: \\
$\bullet$ The minimum exposure time is 1 second, with a stretch goal of 0.1 seconds (OSS-REQ-0291). \\
$\bullet$ The maximum exposure time is not restricted. \\
$\bullet$ However, for exposure times there are other considerations. The minimum exposure time needed to create an image with a PSF that is well-formed enough for difference imaging is a separate question. Changing the exposure time also affects the photometric and astrometric calibrations. Assuming a 1 second exposure can be reduced and calibrated, its detected point sources will span $13 < r < 21$ magnitudes, whereas a 15 second exposure saturates at $r\sim15.8$ mag. A 150 second image would saturate at $r\sim18.3$, perhaps leaving too few stars overlapping with e.g., templates or WFD images, for astrometric and photometric calibrations. Additionally, the impact on CR rejection routines is untested for long exposures.

Camera Rotation: \\
$\bullet$ The OSS contains requirements for rotator capability (OSS-REQ-0301, -0300), but is there a maximum rotation speed, or a limit on the amount of rotation per night (or in a 10-year lifetime), that might constrain the distance between successive visits or the ability to jump between two widely separated fields? \textit{MLG - I heard there might be a lifetime maximum on the rotator, such that if a mini-survey wants to jump between patches that are far apart on the sky repeatedly, that could wear out the mechanism? Can't remember where I heard it.}

Telescope Slews: \\
$\bullet$ \textit{MLG - Obviously, long slews would cause additional overheads and a loss of observing efficiency, but are there other limitations?}

Total Time: \\
$\bullet$ It has been regularly stated that 10\% of the total observing time will be used for special programs (90\% goes to the WFD main survey). This could be increased a bit, but otherwise imposes an upper limit on the total time request of a given Special Program.

Operations Simulations: \\
$\bullet$ Mini-survey observing strategies will have to show that they do not negatively impact the LSST's ability to complete the wide-fast-deep survey; not necessarily in the next call for white papers, but later during the review and revise stage, once the new OpSim runs incorporating proposed mini-surveys have been generated.

\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{Bounds-1} The purpose of this ticket is to request clarification regarding several aspects of the telescope and camera capabilities that might impose boundaries on the diversity of data from Special Programs. First, have the following proposed filter change tolerances been confirmed: (1) the minimum time between filter changes has no restrictions from thermal tolerances, but it is recommended to keep the filter change rate lower than once every 10 minutes in order to minimize overheads; (2) the maximum total number of filter changes is 100000 over 15 years (18/night) and the maximum number of filter swaps in and out of the carousel is 3000 in 15 years (once every two nights). Second, are there any restrictions on the use of the camera rotator or the total amount of slew time? For example, could a Special Program continuously jump between well-separated fields on the sky for an extended amount of time? \end{enumerate}

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
\clearpage
\section{The Current DM Plans for Special Programs Data Processing} \label{sec:dmplans}

As presented in Section \ref{sec:intro}, the intent of LSST DM for processing data from Special Programs is that ``LSST will not write unique algorithms for processing Special Programs data or reprocessing Main Survey data, but LSST will reconfigure the pipelines and generate imaging and catalog products for Special Programs data whenever possible" (\DPDD, \citeds{LSE-163}), and that an additional $\sim$10\% of the LSST computational system has been added and will be dedicated to user processing (of main survey or special programs data).

In this section we review in greater detail the planned ability of DM to process non-standard visit images (Section \ref{ssec:dmplans_NSV}), the intent of DM to incorporate Special Programs data into the pipelines and products for the WFD main survey (Section \ref{ssec:dmplans_WFD}), to reconfigure the pipelines and generate unique sets of data products for each Special Program (Section \ref{ssec:dmplans_reconfig}), and to enable Level 3 processing (Section \ref{ssec:dmplans_L3}).

% % % % % % % % % % % % % % % % % %
\subsection{DM Constraints on Processing Non-Standard Visit Images}\label{ssec:dmplans_NSV}

A standard visit image is considered to be two snap exposures of $\sim15$ second each, and in the WFD main survey sequential visits will be of different fields. As we found in Section \ref{sec:data}, Special Programs are most likely to request observing modes with shorter exposure times and long sequences of visits to the same field. In this section we review DM's planned capability to process such non-standard visit sequences, keeping in mind that DM's processing boundaries might ultimately be defined not by exposure acquisition parameters, but by the image quality parameters such as number of stars with sufficient flux for calibration or the source area density (crowding), and that these boundaries might not be fully constrained until the final performance of DM algorithms as described in the Data Management Applications Design (DMAD, \citeds{LDM-151}) document are characterized (see also Section \ref{ssec:docrev_dmad}).

@@

Images which deviate significantly from the $15$ second duration for the WFD main survey may encounter issues in the instrument signature removal routine, in difference imaging and the correction for differential chromatic refraction, and/or photometric and astrometric calibrations due to a differently sampled set of standard stars per CCD. We consider shorter and longer exposure in turn.

\textbf{Shorter Exposures.} The camera constraint on the minimum supported exposure time is currently 1 second (stretch goal 0.1 seconds). The minimum exposure time for an image to be successfully reduced with Instrument Signature Removal (ISR) will be a DM constraint. Assuming that 1 second exposure can be reduced and calibrated, its detected point sources will span a dynamic range of $r \approx 12.9$ -- $21.0$ magnitudes. A template image built on $15$ second exposures will saturate at $r \approx 15.8$, but this still leaves stars between $15.8$--$21.0$ magnitudes to be used in the PSF-matching (and all other filters have a similarly large overlap). However, in order for an image to be successfully PSF-matched to the template, the PSF must be well formed (i.e., no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (i.e., be smoothly varying on some minimal scale). As a simple demonstration, MLG has made Figure \ref{fig:expt} to show that perhaps exposure times shorter than 2 seconds do not have a well-formed PSF (using the centroid of a 2D Gaussian fit as a proxy for "well-formed"). Another concern is that the Level 1 pipeline would not be able to deliver the required $60$ second latency if the camera produces more than 1 image in 30 seconds, which would disqualify a contribution to the Alert Stream.

@@

\textbf{Longer Exposures.} There is no maximum exposure time specified for an LSST image. Given that the template image will be a stack of at least a year or two of data, processing a $5$--$10$ times deeper single image through the difference imaging pipeline should be fine. However, a $2\times150$ second exposure would saturate at $r \approx 18.3$, and cosmic-ray rejection completeness might suffer (unknown), which could impact the quality of a difference image and the detected sources. With a longer exposure time the $60$ second latency requirement could still be met, but any system qualities that vary on short (but $>30$ second) timescales could inhibit photometric calibration (e.g., tracking).

In conversation with DM-AP team members (Reiss, Findeisen, Connolly, Bo) there has not yet been a study of the safe range of exposure times that will be allowed to contribute to the Level 1 and Alert Stream. One possibly useful study is Chang et al. (2012), "Atmospheric point spread function interpolation for weak lensing in short exposure imaging data". They show that a 15 second exposure contains PSF variability on short spatial scales across a 1 square degree image which, for extragalactic fields with few stars (i.e., but good for weak lensing), is hard to characterize. They present a new software package to do mitigate the effects. Alternatively, we may need to use software packages \texttt{PhoSim} (Peterson et al. 2015; \citep{2015ApJS..218...14P}) or \texttt{ARROYO} \citep{2004SPIE.5497..290B} to at least simply characterize the PSF stability as a function of exposure time.

In conversation with K.-T., exposure times of 10 to 120 seconds will not be a problem for the processing pipeline to accommodate, but the ingest rate of images and the number of exposure per visit are the aspects that could cause issues (discussed below).

Is it conceivable that the LSST would spend time on a Special Program that will acquire short exposures if DM's algorithms cannot be shown to adequately reduce and calibrate them? This is probably best left ot a policy issue (beyond the scope of this document).

\subsubsection{Image Acquisition Rate}

OSS-REQ-0291 requires that short exposures be spaced by 15 seconds, and so DM's ingest rate of exposures should never be more than 1 per 15 seconds, or a visit rate of 1 per 30 seconds. However, \textit{if a higher image acquisition rate is cleared by the camera team}, then DM might encounter some issues such as the minimum time required to save an image to disk, the handling of a processing backup, and/or inconsistently associated \texttt{DIASources}.

\textbf{Minimum Save Time:} If short exposures are taken in an unbuffered sequence, the minimum time between exposures is the shutter-crossing time (1 second) plus the readout time (2 seconds): 3 seconds. This is too rapid an acquisition rate for processing with Level 1, but what if the science goals accept saving them to disk and analyzing them later? \textbf{How long does it take to save an image to disk?}

\textbf{Processing backup:} Are there options for short-term increases in parallel processing power at NCSA? (Such options might be needed anyway to process crowded fields with $>$10k \texttt{DIASources}.) The bandwidth needed to load templates at a faster rate is also a concern, since templates have twice as many pixels, if the data acquistion rate is surveying at twice the WFD main survey rate, that's a $4\times$ additional bandwidth load. However, at NCSA there should be ways to elastically change the amount of processing power available. This is also an issue for crowded fields (below).

\textbf{Unassociated \texttt{DIASources}:} the consistency of \texttt{DIAObjects} suffers if the same patch of sky is observed in quick succession (overlapping regions of short exposures and/or sustained 30 second visits to the same field). In other words, it is not guaranteed that the next string of processing would have access to the latest \texttt{DIAObjects} for association. It is furthermore unclear whether we can flag an Alert as potentially having incomplete source-object associations. This is also an issue for crowded fields (below).

\subsubsection{Exposures per Visit, and Long Sequences of a Single Field}

K.-T. Lim has pointed out that an odd number of exposures is a non-standard visit; two snaps is hardwired into the code. This is baked-in to a configuration so that the pipeline can have a definition of what kind of timing delay constitutes ``late".  Moving away from 2 exposures per visit requires a configuration change to the pipelines, which incurs an overhead (up to 1 minute) -- in fact, K.-T. things that between $10$ and $120$ seconds exposure times can easily be handled by the pipeline (i.e., can be run through ISR using scaled calibration frames), so long as they come in pairs. The real problem is knowing how long the processing should take, and not killing a process that is taking longer because there were 4 snaps in the visit instead of 2. To accommodate non-standard visits requires that the scheduler pass on the information of the number of snaps in the visit (\ref{DMSR-1}). Then the processing pipeline will know to, e.g., not attempt to difference the two snaps in the case were there is an odd number of snaps in a visit. \textit{MLG -- I've heard rumors of a CR regarding alternate standard visits of $1\times30$ seconds, but do not know the status or implications of this.}

>>>>

\documentclass[DM,lsstdoc,toc]{lsstdoc}
\usepackage{graphicx}
\usepackage{url}
\usepackage{latexsym}
\usepackage{color}
% black, blue, brown, cyan, darkgray, gray, green, lightgray, lime, magenta, blue, orange, pink, purple, red, teal, violet, white, yellow.
\usepackage{enumitem}

@@

\setDocRef{DMTN-065}
\date{\today}
\setDocUpstreamLocation{\url{https://github.com/lsst-dm/dmtn-065}}

\setDocAbstract{This document provides an in-depth description of the role of the LSST Project in preparing software and providing computational resources to process the data from Special Programs (deep drilling fields and/or mini-surveys). The main target audience is the LSST Data Management (DM) team, but members of the community who are preparing white papers on science-driven observing strategies may also find this document useful. The potential diversity of data from Special Programs is summarized, including boundaries imposed by technical limitations of the LSST hardware. The capability of the planned Data Management system to processes this diversity of Special Programs data is the main focus of this document. Case studies are provided as examples of how the LSST software and/or user-generated pipelines may be combined to process the data from Special Programs.}

\setDocChangeRecord{%
\addtohist{1}{2017-11-14}{Status: internal working document.}{Melissa Graham}
\addtohist{1}{2018-06-17}{Updated to finalize and issue.}{Melissa Graham}
%\addtohist{2}{yyyy-mm-dd}{Future changes}{Future person}
}

@@

The main LSST science goals will be met by the Wide-Fast-Deep (WFD) Main Survey, but this is expected to be accomplished with $85$--$90$\% of the observing time available over the $10$ year survey. The remaining $10$--$15$\% of the time will be spent on Special Programs: alternative survey areas and/or observing strategies driven by a specific science goal. A call for white papers that provide scientific motivation and observing strategies for the WFD survey and Special Programs, \citeds{Document-28382}, released in June 2018, has a deadline of November 30 2018. It is conceivable that Special Programs might obtain imaging data that are significantly different from the WFD main survey, and/or requires special processing in order to achieve the program's science goals. This document provides an in-depth description of the role of the LSST Project in preparing to process the data from Special Programs (expanding on the general description in Section 6 of \citeds{LSE-163}). The main target audience is the LSST Data Management (DM) team, but members of the science community who are preparing white paper proposals may also find this document useful.

{\bf Special Programs Terminology -- } The WFD main survey will survey the sky with a series of {\bf visits}, and each field will be visited $>800$ times over $10$ years. A {\bf standard visit} is composed of $2\times15$ second exposures (commonly referred to as ``snaps") and an {\bf alternative standard visit} is composed of a single $30$ second exposure. A {\bf non-standard visit} is any other exposure time(s) or number of snaps. Special Programs are typically divided into two types: {\bf Deep Drilling}, a single pointing for which many exposures are obtained in a relatively short amount of time (e.g., $>2\times$ as many visits in six months as the WFD will obtain in 10 years); and {\bf Mini-Surveys}, which refer to either new sky areas observed with a WFD-like survey, or sky areas within the WFD but observed with a specialized strategy.

{\bf The LSST Project's Role in Processing Special Programs Data -- } The LSST Project will not take formal responsibility for specialized data reduction algorithms needed to process data, including images taken in non-standard modes. The term "specialized algorithms" refers to software that is not already within scope of the LSST Data Management (DM) science pipelines, and may include, for example: difference imaging for short exposures in which the PSF is not well-formed, shift-and-stack for faint moving objects, or any software with computational needs that significantly surpass the processing budget per image (compared to the processing of a WFD image). The Project will incorporate Special Programs data into the Prompt and/or Data Release processing pipelines and data products of the WFD Main Survey, such as {\tt Alerts}, {\tt CoAdds}, or {\tt Source} and {\tt Object} catalogs (with appropriate flags; LSST data products are described in \citeds{LSE-163}), whenever this (1) can be accomplished with existing software, and (2) is scientifically beneficial to that data product. The Project will also reconfigure its pipelines to generate separate imaging and catalog data products for Special Programs, whenever this can be accomplished with existing software. Finally, the Project will enable user-generated processing via the Science Platform (\citeds{LSE-319}), which will provide software tools and computational resources for (re)processing LSST data.

{\bf Document Overview -- } The purpose of this document is to assess the potential diversity of imaging data that might be obtained by Special Programs (Section \ref{sec:data}), and to explore and clarify whether -- and to what extent -- DM's planned pipelines and user services will be able to handle this diversity of data (Section \ref{sec:dmplans}). As described in the call for white papers on cadence optimization (\citeds{Document-28382}), Special Programs that may require specialized or computationally-intense algorithms to meet their science goals are required to describe how these processing needs will be met. Sections \ref{sec:data} and \ref{sec:dmplans} have been designed to help white paper authors figure out whether their proposed Science Program(s) require processing that would be considered out of DM's scope, and in Section \ref{sec:SPCS} we have provided a series of science-driven case studies as examples.

% In Appendix \ref{sec:docrev}, we endeavor to ensure that the DM's plans for Special Programs processing are appropriately expressed in all of the major Project documents and that the computational system and human work-hours needed to accomplish Special Programs processing have been accurately evaluated. This content will only be of interest to the DM team.

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
\clearpage
\section{The Potential Diversity of Special Programs Data} \label{sec:data}

To define the extent to which Data Management will be able to process data from Special Programs, a comprehensive understanding of the potential diversity of data from Special Programs -- compared to the WFD's sky survey of standard visits -- is needed.

To build up this comprehensive understanding, in Section \ref{ssec:data_bounds} we considered any technical limitations that the facility and instrumentation will or might place on the data. We find that the hardware imposes few boundaries on how data can be obtained, but that a high number of filter changes and/or long slews are inefficient due to their large overheads. The minimum exposure time is $1$ second (stretch goal: $0.1$ seconds), but there is currently a technical boundary that limits the readout rate to $1$ every $15$ seconds.

Next, we considered the Special Programs that have been openly discussed so far in the Science Community (Appendix \ref{sec:data_prev}). Based on this consideration, most of the proposed Special Programs are likely to use visits that are similar to the WFD Main Survey, but some will require exposures that are significantly shorter, or that are obtained with a bright sky background during twilight. The cadence and patterns may also differ from the WFD main survey, such as long series of exposures obtained of the same field (i.e., deep drilling), or a strategy optimized to find very fast-moving objects. In addition, images of very crowded fields in the Galactic Plane may be included (at all or more often than the WFD Main Survey). It does not appear that any of the previously proposed Special Programs violate the hardware-imposed boundaries discussed in Section \ref{ssec:data_bounds}.

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
\subsection{Hardware Boundaries}\label{ssec:data_bounds}

Here we consider the technical boundaries on the diversity of data products that are expected to (or may) be imposed by limitations from the camera, telescope, and/or site (boundaries imposed by DM processing capabilities are considered in Section \ref{sec:dmplans}).

{\bf Filter Changes -- } The maximum time for filter change is $120$ seconds: $30$ seconds for the telescope to reorient the camera to its nominal zero angle position on the rotator, and $90$ seconds to the camera subsystem for executing the change (OSS-REQ-0293; \citeds{LSE-30}). Assuming that most Special Programs would be designed to keep overheads $<100\%$ and would be using standard $30$ second visits, the filter change time indicates that it is likely that at least $4 exposures$ in a given filter would be obtained between filter changes -- but this is not actually a technical boundary.

{\bf Filter Carousel Loads -- } As described in \citeds{Document-28382}, the filter carousel can hold five of the six LSST filters at a time, and filter loads are done in the day. The system is designed to support $3000$ loads and $100000$ filter changes in $15$ years, which is an average of $17$ changes per night (after accounting for filter changes during calibrations). Individual filters will support $30000$ changes in $15$ years. Based on these technical boundaries, we know that there will never be data in more than five filters in a given night.

{\bf Exposure Times -- } The minimum exposure time is $1$ second, with a stretch goal of $0.1$ seconds (OSS-REQ-0291; \citeds{LSE-30}). The maximum exposure time is not restricted. The readout time is $2$ seconds, and would be significant overhead on short exposures. Images with exposure times $<15$ seconds may still have to be separated by $15$ seconds for thermal tolerance; i.e., that the minimum readout rate is one image every $15$ seconds, regardless of exposure time (OSS-REQ-0291; \citeds{LSE-30}). We therefore consider the $15$ interval between images a technical boundary on the potential diversity of data products.
% $\bullet$ However, for exposure times there are other considerations. The minimum exposure time needed to create an image with a PSF that is well-formed enough for difference imaging is a separate question. Changing the exposure time also affects the photometric and astrometric calibrations. Assuming a 1 second exposure can be reduced and calibrated, its detected point sources will span $13 < r < 21$ magnitudes, whereas a 15 second exposure saturates at $r\sim15.8$ mag. A 150 second image would saturate at $r\sim18.3$, perhaps leaving too few stars overlapping with e.g., templates or WFD images, for astrometric and photometric calibrations. Additionally, the impact on CR rejection routines is untested for long exposures.

{\bf Telescope Slew -- } As described in \citeds{Document-28382}, large slews would have considerable overheads, but there are no technical boundaries on the size of a single slew or the accrued slew distance.

{\bf Camera Rotation -- } There requirements on the rotator's capabilities do not set any limits on the per-night or total lifetime rotation (OSS-REQ-0301, -0300; \citeds{LSE-30}) which might put boundaries on the distance between successive visits or the ability to jump between two widely separated fields. JIRA ticket DM-12573 is currently open and asking for clarification from the camera team on this. Until then, we assume there are no technical boundaries imposed by camera rotation constraints on the potential diversity of data products.

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
\clearpage
\section{Processing Special Programs Data with LSST DM Pipelines} \label{sec:dmplans}

In this section we go into greater detail regarding the Projects' role in processing data from Special Programs, as introduced in Section \ref{sec:intro}. We discuss the ability of the planned DM pipelines to: process diverse imaging data that are unlike the WFD's standard visits in Section \ref{ssec:dmplans_NSV}; incorporate Special Programs data into the Prompt and Data Release pipelines and data products for the WFD main survey in Section \ref{ssec:dmplans_WFD}; reconfigure the pipelines and generate unique sets of data products for each Special Program in Section \ref{ssec:dmplans_reconfig}; and enable user-generated pipelines and data products in Section \ref{ssec:dmplans_user}.

% % % % % % % % % % % % % % % % % %
\subsection{Processing Diverse Imaging Data Unlike the WFD Main Survey}\label{ssec:dmplans_NSV}

As described in Section \ref{sec:data}, Special Programs will be able -- and likely -- to request observing modes with shorter (or longer) exposure times, long sequences of visits to the same field, and/or imaging of very crowded fields. We review the capability of DM's planned pipelines to process such each of these kinds of diverse data, keeping in mind that the processing boundaries might ultimately be defined not by what is technically possible, but by the resulting image quality parameters such as the number of stars with sufficient flux for photometric calibration or the source area density (crowding). Furthermore, that these boundaries imposed by the data quality might not be constrained until the final performance of DM's algorithms, as described in the Data Management Applications Design (\citeds{LDM-151}) document, is fully characterized.

@@

Images which deviate significantly from the $15$ second duration for the WFD main survey may encounter issues in the instrument signature removal routine, in the correction for differential chromatic refraction, in the difference imaging analysis pipeline, and/or in the photometric and astrometric calibrations due to a differently sampled set of standard stars per CCD. We discuss shorter and longer exposure in turn.

\textbf{Shorter Exposures.} The camera constraint on the minimum supported exposure time is currently 1 second (stretch goal 0.1 seconds). The minimum exposure time for an image to be successfully reduced with Instrument Signature Removal (ISR) is under consideration (see JIRA ticket DM-12574). Assuming that 1 second exposure can be reduced and calibrated, its detected point sources will span a dynamic range of $r \approx 12.9$ -- $21.0$ magnitudes. A template image built on $15$ second exposures will saturate at $r \approx 15.8$, but this still leaves stars between $15.8$--$21.0$ magnitudes to be used in the PSF-matching (and all other filters have a similarly large overlap). However, in order for an image to be successfully PSF-matched to the template, the PSF must be well formed (no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (be smoothly varying on some minimal scale). As a simple demonstration, Figure \ref{fig:expt} shows that perhaps exposure times shorter than $2$ seconds do not have a well-formed PSF (using the centroid of a 2D Gaussian fit as a proxy for "well-formed"). The minimum exposure time for an image to be successfully processed by the Difference Imaging Analysis (DIA) pipeline is currently under consideration (see JIRA ticket DM-12574).

@@

\textbf{Longer Exposures.} There is no maximum exposure time specified for an LSST image. Given that the template image will be a stack of at least a year or two of data, processing a $5$--$10$ times deeper single image through the difference imaging pipeline should be fine. However, a $2\times150$ second exposure would saturate at $r \approx 18.3$, and cosmic-ray rejection completeness might suffer (unknown), which could impact the quality of a difference image and the detected sources. Additionally, any system qualities that vary on short (but $>30$ second) timescales could inhibit photometric calibration (e.g., tracking). A potential maximum exposure time from a processing perspective is currently under consideration (see JIRA ticket DM-12574).

% In conversation with DM-AP team members (Reiss, Findeisen, Connolly, Bo) there has not yet been a study of the safe range of exposure times that will be allowed to contribute to the Level 1 and Alert Stream. One possibly useful study is Chang et al. (2012), "Atmospheric point spread function interpolation for weak lensing in short exposure imaging data". They show that a 15 second exposure contains PSF variability on short spatial scales across a 1 square degree image which, for extragalactic fields with few stars (i.e., but good for weak lensing), is hard to characterize. They present a new software package to do mitigate the effects. Alternatively, we may need to use software packages \texttt{PhoSim} (Peterson et al. 2015; \citep{2015ApJS..218...14P}) or \texttt{ARROYO} \citep{2004SPIE.5497..290B} to at least simply characterize the PSF stability as a function of exposure time.
% In conversation with K.-T., exposure times of 10 to 120 seconds will not be a problem for the processing pipeline to accommodate, but the ingest rate of images and the number of exposure per visit are the aspects that could cause issues (discussed below).
% Is it conceivable that the LSST would spend time on a Special Program that will acquire short exposures if DM's algorithms cannot be shown to adequately reduce and calibrate them? This is probably best left ot a policy issue (beyond the scope of this document).

\subsubsection{Number of Exposures per Visit (Long Sequences of a Single Field)}

There is no processing constraint on the number of consecutive exposures that could be obtained of a single field. From a DM perspective, it would be best if these exposures were packaged into visits of no more than 2 exposures per visit, to minimize the need to reconfigure of the pipelines, and because the camera only ``clears" between visits.

% K.-T. Lim has pointed out that an odd number of exposures is a non-standard visit; two snaps is hardwired into the code. This is baked-in to a configuration so that the pipeline can have a definition of what kind of timing delay constitutes ``late".  Moving away from 2 exposures per visit requires a configuration change to the pipelines, which incurs an overhead (up to 1 minute) -- in fact, K.-T. things that between $10$ and $120$ seconds exposure times can easily be handled by the pipeline (i.e., can be run through ISR using scaled calibration frames), so long as they come in pairs. The real problem is knowing how long the processing should take, and not killing a process that is taking longer because there were 4 snaps in the visit instead of 2. To accommodate non-standard visits requires that the scheduler pass on the information of the number of snaps in the visit (\ref{DMSR-1}). Then the processing pipeline will know to, e.g., not attempt to difference the two snaps in the case were there is an odd number of snaps in a visit. \textit{MLG -- I've heard rumors of a CR regarding alternate standard visits of $1\times30$ seconds, but do not know the status or implications of this.}

% K.-T. has also pointed out that currently, a deep drilling field would be interpreted as a single visit of 50 exposures by the scheduler. One implication of this is that since the camera only ``clears" prior to a new visit, it would not do this for the entire 50-exposure sequence. The processing pipeline would need to know how to divide this sequence up into visits. As there is no current requirement for DM to receive the information that the scheduler is about to do a 50 exposure visit, we need \ref{DMSR-1} to add the proposal ID and the number of exposures per visit to the meta data, and then it should be OK for DM to parse this visit information in the reduction pipeline.

\subsubsection{Images in Very Crowded Fields}

The LSST pipelines' performance in crowded fields is documented in \citeds{DMTN-077}, which finds that, e.g., in Galactic Plane regions with a source density of $500000$ sources per square degree, the completeness drops to 50\% at $20.2$ magnitudes. The slide deck at \citeds{Document-27962} also describes DM's plans for processing crowded fields. These may or may not be appropriate for Special Programs data, depending on the science goals.

% % % % % % % % % % % % % % % % % %
\subsection{Including Special Programs Data in the WFD Main Survey's Data Products}\label{ssec:dmplans_WFD}

As described in Section \ref{sec:intro}, the Project may incorporate Special Programs data into the WFD main survey's pipelines and data products whenever this is (1) possible and (2) scientifically beneficial. This will likely be at the discretion of the data quality assessment team during Operations. In the following three sections we project when and how Special Programs data might be incorporated into the pipelines and data products of the Prompt pipeline and the Alert stream (Section \ref{ssec:dmplans_prompt}) and the annual Data Release pipeline (Section \ref{ssec:dmplans_drp}).

\subsubsection{The Prompt Pipeline and Alert Generation}\label{ssec:dmplans_prompt}
